Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-05 ind_files/03_BD_processes.tex
Add explicit nod to Rogers in 3.1

Raymond Pearl, one of the pioneers of mathematical ecology as a discipline, wrote in 1927 that ``What we want to know is how the biological forces of natality and mortality are so integrated and correlated in their action as to lead to a final result in size of population which follows this particular curve rather than some other one'' \citep{pearl_growth_1976}. Pearl realized 95 years ago that population dynamics must be ultimately explained by the mechanistic processes of birth and death. Today, there are mounting calls for more mechanistic models of evolution that are grounded in these fundamental processes of birth and death \citep{geritz_mathematical_2012,doebeli_towards_2017}. Ecologists also increasingly recognize that incorporating stochasticity is vital to developing more realistic ecological models \citep{hastings_transients_2004, coulson_skeletons_2004, boettiger_noise_2018, shoemaker_integrating_2020,schreiber_does_2022} and does more than `add noise' to deterministic expectations. Individual-based models, where ecological rules are specified at the level of the individual, are a powerful mathematical tool for mechanistic descriptions of stochastic population dynamics \citep{black_stochastic_2012}. Birth-death processes are a very general class of stochastic processes that can be used to capture a wide range of eco-evolutionary processes. `System-size expansions' and their subsequent analysis using the `weak noise approximation' are common tools for analyzing stochastic birth-death processes that are well-known in the statistical physics and applied mathematics communities \citep{gardiner_stochastic_2009}. However, these tools remain relatively underappreciated in ecology, despite being relatively easy to understand and extremely well-motivated in scenarios germane to ecology and evolution. Here, I present a formulation of population dynamics constructed from first principles grounded in birth-death processes. To facilitate readership by a broad audience, only passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability are assumed. Familiarity with stochastic calculus is helpful for some sections but is not required, and I provide a brief introduction below. I begin by introducing birth-death processes, SDEs, and the Fokker-Planck equation in a non-technical manner, and providing the intuition for system-size expansions and weak noise analysis in ecological systems. In section \ref{sec_1D_processes}, I show how fluctuating population size of populations of identical individuals can be tracked through a one-dimensional birth-death process. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process. As an example, I illustrate the complete process for a stochastic version of the logistic equation. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, illustrate the system size expansion and the weak noise approximation. Under mild assumptions, we show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations in evolutionary game theory. I also show that the mean value of the trait in the population changes according to the Price equation in the deterministic limit. Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck equation, in terms of functional derivatives. Under mild assumptions, we show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. We also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation.  Sections \ref{sec_1D_processes} and \ref{sec_nD_processes} apply techniques that are well-known in physics, as applied to population biology. In contrast, chapter \ref{chap_infD_processes}, to the best of my knowledge, is entirely original. Chapter \ref{chap_infD_processes} also presents a simple heuristic derivation of quantitative genetics models and adaptive dynamics from stochastic first principles that is much simpler than the rigorous mathematical derivations grounded in measure theory and martingale/markov theory that are currently standard reference in theoretical ecology \citep{champagnat_individual_2008}. I illustrate the utility of all this abstract formalism through examples in Chapter \ref{chap_examples}.

>>>>

Raymond Pearl, one of the pioneers of mathematical ecology as a discipline, wrote in 1927 that ``What we want to know is how the biological forces of natality and mortality are so integrated and correlated in their action as to lead to a final result in size of population which follows this particular curve rather than some other one'' \citep{pearl_growth_1976}. Pearl realized 95 years ago that population dynamics must be ultimately explained by the mechanistic processes of birth and death. Today, there are mounting calls for more mechanistic models of evolution that are grounded in these fundamental processes of birth and death \citep{geritz_mathematical_2012,doebeli_towards_2017}. Ecologists also increasingly recognize that incorporating stochasticity is vital to developing more realistic ecological models \citep{hastings_transients_2004, coulson_skeletons_2004, boettiger_noise_2018, shoemaker_integrating_2020,schreiber_does_2022} and does more than `add noise' to deterministic expectations. Individual-based models, where ecological rules are specified at the level of the individual, are a powerful mathematical tool for mechanistic descriptions of stochastic population dynamics \citep{black_stochastic_2012}. Birth-death processes are a very general class of stochastic processes that can be used to capture a wide range of eco-evolutionary processes. `System-size expansions' and their subsequent analysis using the `weak noise approximation' are common tools for analyzing stochastic birth-death processes that are well-known in the statistical physics and applied mathematics communities \citep{gardiner_stochastic_2009}. However, these tools remain relatively underappreciated in ecology, despite being relatively easy to understand and extremely well-motivated in scenarios germane to ecology and evolution. Here, I present a formulation of population dynamics constructed from first principles grounded in birth-death processes. To facilitate readership by a broad audience, only passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability are assumed. Familiarity with stochastic calculus is helpful for some sections but is not required, and I provide a brief introduction below. I begin by introducing birth-death processes, SDEs, and the Fokker-Planck equation in a non-technical manner, and providing the intuition for system-size expansions and weak noise analysis in ecological systems. In section \ref{sec_1D_processes}, I show how fluctuating population size of populations of identical individuals can be tracked through a one-dimensional birth-death process. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process. As an example, I illustrate the complete process for a stochastic version of the logistic equation. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, illustrate the system size expansion and the weak noise approximation. Under mild assumptions, we show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations in evolutionary game theory. I also show that the mean value of the trait in the population changes according to the Price equation in the deterministic limit. Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck equation, in terms of functional derivatives. Under mild assumptions, we show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. We also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation.  Sections \ref{sec_1D_processes} and \ref{sec_nD_processes} apply techniques that are well-known in physics, as applied to population biology. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues \citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, is entirely original. Chapter \ref{chap_infD_processes} also presents a simple heuristic derivation of quantitative genetics models and adaptive dynamics from stochastic first principles that is much simpler than the rigorous mathematical derivations grounded in measure theory and martingale/markov theory that are currently standard reference in theoretical ecology \citep{champagnat_individual_2008}. I illustrate the utility of all this abstract formalism through examples in Chapter \ref{chap_examples}.

### 2023-02-05 main.tex
Separate results from discussion
Separate field eqns into separate chapter
Modify TOC to include list of tables & figures
Add to discussion in chap 7
Add explicit nod to Rogers in 3.1

%%%%%%%%%%%%
% Document %
%%%%%%%%%%%%
%\setcounter{chapter}{-1}

\title{Eco-evolutionary dynamics of finite populations from first principles}

\author{Ananda Shikhara Bhat}
\coordinator{Coordinator} \supervisor{Vishwesha Guttal}\cosupervisor{Rohini Balakrishnan} \sdesignation{}
\department{Centre for Ecological Sciences, Indian Institute of Science} \reader{Sutirth Dey}
%\reader{Reader 2}
\dedication{This thesis is dedicated to ?}
\graduationyear{2023}
\academicyear{2022-2023}
\graduationmonth{April}

\thesisabstract{
Some central questions of ecology and evolution ask how many distinct variants of an entity can emerge and coexist in sympatry. Examples include the study of standing genetic variation (alleles), polymorphisms (genotypes/phenotypes), and sympatric speciation (species). Historically, these questions have often been studied through phenomenological models formulated for infinitely large populations. Stochastic individual-based models (IbMs), where mechanistic rules are specified for each individual, are more realistic but are rarely studied analytically.\\
\\
In this thesis, I show how ideas from statistical physics can be used to analytically formulate and study IbMs from biological first principles. Starting from a stochastic ‘birth-death process’, I show how density dependence of the population dynamics can be used to obtain a continuous approximation of the system as a stochastic differential equation (SDE). I show that well-known ‘fundamental theorems’ of evolution, such as the Price equation, are recovered in the infinite population limit. My stochastic formulation reveals a new directional evolutionary force in finite populations that lends itself to a simple biological interpretation and has no analogue in infinitely large populations. As an application, I show how this formalism can be used to study cooperation in finite populations. Finally, I extend this theory to study quantitative traits through a stochastic field equation, which can be thought of as an infinite-dimensional SDE. In the infinite population limit, I recover well-known frameworks such as Kimura's continuum-of-alleles and adaptive dynamics. My work thus highlights the connections between various equations in ecology and evolution by revealing their common stochastic underpinnings.
}

@@

\pagestyle{plain}
\thesisfront
\listoftables
\listoffigures

@@

%Implement the fancy headers for everything from here on out
\pagestyle{fancy}
%\fancyhead{}
%\fancyhead[RO]{Chapter \thechapter}
%\fancyhead[LE]{\mytitle}

\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[RE]{\rule[-2ex]{0pt}{4ex}\footnotesize\itshape\nouppercase{\leftmark}}
\fancyhead[LO]{\rule[-2ex]{0pt}{4ex}\footnotesize\itshape\nouppercase{\rightmark}}
\setlength{\headheight}{18pt} % as requested by fancyhdf
\renewcommand{\headrulewidth}{0pt}

@@

\part{Theory \& Applications}
\chapter{Population dynamics from stochastic first principles}\label{chap_BD}
\input{ind_files/03_BD_processes.tex}
\chapter{Examples and Applications}\label{chap_examples}
\input{ind_files/04_examples}
% \chapter{}
% \input{ind_files/chapter_05}
% \chapter{} <-- uncomment and add as required
% \input{ind_files/chapter_06}

\begin{appendices}
\chapter{Deriving the Fokker-Planck equations for It\^{o} SDEs}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\end{appendices}

>>>>

%%%%%%%%%%%%
% Document %
%%%%%%%%%%%%

\title{Eco-evolutionary dynamics of finite populations from first principles}
\author{Ananda Shikhara Bhat}

%Details of supervisors and experts
\supervisor{Vishwesha Guttal}
\cosupervisor{Rohini Balakrishnan}
\department{Centre for Ecological Sciences, Indian Institute of Science} \reader{Sutirth Dey} %Expert

\dedication{This thesis is dedicated to ?}

%Details of graduation date
\graduationyear{2023}
\academicyear{2022-2023}
\graduationmonth{May}

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Some central questions of ecology and evolution ask how many distinct variants of an entity can emerge and coexist in sympatry. Examples include the study of standing genetic variation (alleles), polymorphisms (genotypes/phenotypes), and sympatric speciation (species). Historically, these questions have often been studied through phenomenological models formulated for infinitely large populations. Stochastic individual-based models (IbMs), where mechanistic rules are specified for each individual, are more realistic but are rarely studied analytically.\\
\\
In this thesis, I show how ideas from statistical physics can be used to analytically formulate and study IbMs from biological first principles. Starting from a stochastic ‘birth-death process’, I show how density dependence of the population dynamics can be used to obtain a continuous approximation of the system as a stochastic differential equation (SDE). I show that well-known ‘fundamental theorems’ of evolution, such as the Price equation, are recovered in the infinite population limit. My stochastic formulation reveals a new directional evolutionary force in finite populations that lends itself to a simple biological interpretation and has no analogue in infinitely large populations. As an application, I show how this formalism can be used to study cooperation in finite populations. Finally, I extend this theory to study quantitative traits through a stochastic field equation, which can be thought of as an infinite-dimensional SDE. In the infinite population limit, I recover well-known frameworks such as Kimura's continuum-of-alleles and adaptive dynamics. My work thus highlights the connections between various equations in ecology and evolution by revealing their common stochastic underpinnings.
}

@@

\pagestyle{plain}
\thesisfront
\listoftables
\addcontentsline{toc}{chapter}{List of Tables}
\listoffigures
\addcontentsline{toc}{chapter}{List of Figures}

@@

%Implement the fancy headers for everything from here on out
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{\thepage}
\fancyhead[RE]{\rule[-2ex]{0pt}{4ex}\footnotesize\itshape\nouppercase{\leftmark}}
\fancyhead[LO]{\rule[-2ex]{0pt}{4ex}\footnotesize\itshape\nouppercase{\rightmark}}
\setlength{\headheight}{18pt} % as requested by fancyhdf
\renewcommand{\headrulewidth}{0pt}

@@

\part{Theory \& Applications}
\chapter{Population dynamics from stochastic first principles}\label{chap_BD}
\input{ind_files/03_BD_processes}
\chapter{Stochastic field equations for the evolution of quantitative traits}\label{chap_infD_processes}
\input{ind_files/04_field_eqns}
\chapter{Examples and Applications}\label{chap_examples}
\input{ind_files/05_examples}

\part{Discussion \& Outlook}
\chapter{A unified view of population dynamics}\label{chap_unification}
\input{ind_files/06_unification_discussion}
\chapter{Towards a stochastic evolutionary theory}\label{chap_stochasticity}
\input{ind_files/07_stochastic_discussion}
% \chapter{}
% \input{ind_files/chapter_05}
% \chapter{} <-- uncomment and add as required
% \input{ind_files/chapter_06}

%You need the below commands to elevate the appendix to 'Part' level in the TOC
%from https://tex.stackexchange.com/a/286452
\cleardoublepage
\phantomsection
\appendix
\addcontentsline{toc}{part}{Appendices}

%make sure the title is 'Appendix' and not 'Chapter'
\renewcommand{\chaptername}{Appendix} 
\chapter{Deriving the Fokker-Planck equations for It\^{o} SDEs}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
